Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-08-02 revision/review_response.Rnw

{\it We thank the editor and reviewer for providing thoughtful comments on our manuscript.  Responses to these comments are shown in italics.  Page and paragraph numbers refer to the revised manuscript.}

\Bigtxt{Editor's comments:}

The authors executed the work on parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern Gulf of Mexico. The authors did not provide proper conceptual diagram of the model.

{\it The previous conceptual diagram was for CGEM that included both the biogeochemical components of FishTank and hydrodynamic components.  A new figure was created that inludes only the FishTank components.  }

@@

lines 24 - 39 page 3. I think the reference Levins (1966) is extremely old and must be removed.

{\it This citation was referenced in Ganju et al. 2016, from which this information was obtained.  We have not removed the citation because we feel that age does not make a publication inadequate.  However, we have added the citation to Ganju et al. 2016 as a more current reference.}

It is interesting to see that the authors constantly use the term ``precision'' to denote goodness of fit. In general, precision is used to denote the number of digits of a number. Note that a number can be precise but inaccurate. I suggest the authors to revise the text and replace ``precision'' with a better term to denote goodness of fit. e.g. level of agreement between observations and simulations.

{\it We have found that precision is commonly used in the literature to describe goodness of fit, although we understand the distinction made by the reviewer.  Considering an example where observed data represent a biased sample of an environmental process (e.g., infrequent or unevent sampling of an event that varies over time), model calibration to these data can only possibly achieve precision.  Although practitioners would not knowingly use observed data that are biased for model calibration, we are certain that many ecological datasets are incomplete representations of actual processes.  We argue then that model calibration in many cases is better described as seeking precision, although in practice it is stated as predictive accuracy given an incomplete knowledge of the observed data relative to the true environmental process. Regardless, we have replaced precision with ``goodness of fit'' or ``fit'' as a more generic term.}

@@

Line 9 page 10. Selection or Selected?

{\it Changed to: `Heurestics for parameter selection also recognized...'}

Line 4 page 12. what are the default parameter conditions?. You should include a table with the default parameter values and parameter values under 50\% modified conditions.

{\it The starting parameter values were added to Tables 1, S1-S7.  We have not added the 50\% increase in each parameter as this is a simple calculation and it clutters the tabular information.}

### 2017-08-02 ident_manu.Rnw
parameter values to tables

% aliased citations
\defcitealias{LehrterIP}{Lehrter et al. in press}

@@

\author{Lisa L. Lowe}
\address{Lockheed Martin IS \& GS - Civil supporting the USEPA, Research Triangle Park, NC 27709}
\ead{lowe.lisa@epa.gov}

@@

Hypoxia formation in bottom waters of coastal oceans occurs primarily from excess nutrient inputs from land-based sources \citep{Justic87,Diaz95,Howarth96}.  These events are detrimental to aquatic organisms and have significant negative effects on economic resources derived from coastal ecosystems \citep{Lipton03,Diaz11}.  An understanding of the biological, physical, and chemical processes that influence hypoxic areas is a critical concern for mitigating and preventing these negative impacts.  Numerical ecosystem models are important tools that synthesize knowledge of ecosystem processes that contribute to hypoxia formation and for predicting the effects of proposed management activities or future scenarios \citep{Scavia04,Hagy07,Camacho14a,Pauer16}.  Unlike statistical models with more generic structures, simulation and process-based models include explicit descriptions of relevant processes that are constrained by empirical or observational data relevant to the system of interest \citep[e.g.][]{Omlin01b,Eldridge10}.  These models are often coupled with hydrodynamic grids to provide spatially-explicit representations of patterns in three dimensions \citep{Warner05,Dortch07,Zhao10,Ganju16}. Combined hydrodynamic and bio-geo-chemical models have been developed specifically to describe hypoxic conditions on the \ac{lcs} in the northern \ac{gom} (\citealt{Fennel13,Obenour15,Pauer16}, \citetalias{LehrterIP}).  This area drains a significant portion of the continental United States through the \ac{marb} and is the second largest hypoxic area in the world \citep{Rabalais02}.  Understanding processes that contribute to the frequency and duration of hypoxic events remains a critical research goal for the region.

@@

Hypoxic events, defined  as $<$2 mg L$^{-1}$ of \ac{do} ($<$ 64 mmol m$^{-3}$), occur seasonally in bottom waters in the northern \ac{gom}.  The hypoxic area averages 15,540 km$^2$ annually (1993-2015) with minimum concentrations observed from late spring to early fall.  Seasonal variation is strongly related to carbon and nutrient export from the \ac{marb} \citep{Lohrenz08,Bianchi10}, whereas hydrologic variation, currents, and wind patterns can affect vertical salinity gradients that contribute to hypoxia formation \citep{Wiseman97,Paerl98,Obenour15}. The \ac{cgem} was developed to describe hypoxia dynamics on the \ac{lcs} and includes elements from the Navy Coastal Ocean Model \citep{Martin00} for hydrodynamics and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep[\cref{fig:fishmod},][]{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf.  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.

The core unit of \ac{cgem} is FishTank, a \ac{zerod} model that implements the biogeochemical equations in \citet{Eldridge10} and does not include any advection, mixing, or sediment diagenesis. Although FishTank was developed for specific application in \ac{cgem}, it can easily be applied to other hydrodynamic grids. Results are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups given nutrient uptake rates, organic matter inputs and losses, inherent optical properties, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIP}). A total of 108 equations are estimated at each time step to return values for each of the 36 state variables described by the model.  In addition to the initial conditions, 251 parameter values for each of the equations are also supplied at model execution. Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.

The sensitivity of state variables to perturbations of all relevant parameters for the 108 equations was estimated using a five minute timestep with daily output from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006. Irrelevant parameters were removed for several reasons; parameters were not relevant for the \ac{zerod} model (i.e., hydrodynamic parameters), were considered physical constants, or had no effect given initial conditions.  Additionally, FishTank includes six phytoplankton and two zooplankton groups to add complexity in community structure and foodweb dynamics. To remove obvious redundancies, the sensitivity analyses were conducted using only one phytoplankton and one zooplankton group.  The final set that was evaluated included \Sexpr{length(p1z1pars)} parameters that were further grouped into one of six categories based on applicable biogeochemical components of the model: optics ($n = $ \Sexpr{nparms$Optics} parameters), organic matter (\Sexpr{nparms$`Organic Matter`}), phytoplankton (\Sexpr{nparms$Phytoplankton}), temperature (\Sexpr{nparms$Temperature}), and zooplankton (\Sexpr{nparms$Zooplankton}).  A full description of the model parameters is available as an appendix in \citetalias{LehrterIP}.

>>>>

@@

\author{Lisa L. Lowe}
\address{CSRA, Inc. supporting the USEPA, Research Triangle Park, NC 27709}
\ead{lowe.lisa@epa.gov}

@@

Hypoxia formation in bottom waters of coastal oceans occurs primarily from excess nutrient inputs from land-based sources \citep{Justic87,Diaz95,Howarth96}.  These events are detrimental to aquatic organisms and have significant negative effects on economic resources derived from coastal ecosystems \citep{Lipton03,Diaz11}.  An understanding of the biological, physical, and chemical processes that influence hypoxic areas is a critical concern for mitigating and preventing these negative impacts.  Numerical ecosystem models are important tools that synthesize knowledge of ecosystem processes that contribute to hypoxia formation and for predicting the effects of proposed management activities or future scenarios \citep{Scavia04,Hagy07,Camacho14a,Pauer16}.  Unlike statistical models with more generic structures, simulation and process-based models include explicit descriptions of relevant processes that are constrained by empirical or observational data relevant to the system of interest \citep[e.g.][]{Omlin01b,Eldridge10}.  These models are often coupled with hydrodynamic grids to provide spatially-explicit representations of patterns in three dimensions \citep{Warner05,Dortch07,Zhao10,Ganju16}. Combined hydrodynamic and bio-geo-chemical models have been developed specifically to describe hypoxic conditions on the \ac{lcs} in the northern \ac{gom} \citep{Fennel13,Obenour15,Pauer16,Lehrter17}.  This area drains a significant portion of the continental United States through the \ac{marb} and is the second largest hypoxic area in the world \citep{Rabalais02}.  Understanding processes that contribute to the frequency and duration of hypoxic events remains a critical research goal for the region.

@@

Hypoxic events, defined  as $<$2 mg L$^{-1}$ of \ac{do} ($<$ 64 mmol m$^{-3}$), occur seasonally in bottom waters in the northern \ac{gom}.  The hypoxic area averages 15,540 km$^2$ annually (1993-2015) with minimum concentrations observed from late spring to early fall.  Seasonal variation is strongly related to carbon and nutrient export from the \ac{marb} \citep{Lohrenz08,Bianchi10}, whereas hydrologic variation, currents, and wind patterns can affect vertical salinity gradients that contribute to hypoxia formation \citep{Wiseman97,Paerl98,Obenour15}. The \ac{cgem} was developed to describe hypoxia dynamics on the \ac{lcs} and includes elements from the Navy Coastal Ocean Model \citep{Martin00} for hydrodynamics and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf.  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.

The core unit of \ac{cgem} is FishTank, a \ac{zerod} model that implements the biogeochemical equations in \citet{Eldridge10} and does not include any advection, mixing, or sediment diagenesis (\cref{fig:fishmod}). A full description of the model structure, equations, and parameters is described in \citet{Lehrter17}. Although FishTank was developed for specific application in \ac{cgem}, it can easily be applied to other hydrodynamic grids. Results are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups given nutrient uptake rates, organic matter inputs and losses, inherent optical properties, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citealt{Lehrter17}). A total of 108 equations are estimated at each time step to return values for each of the 36 state variables described by the model.  In addition to the initial conditions, 251 parameter values for each of the equations are also supplied at model execution. Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.

The sensitivity of state variables to perturbations of all relevant parameters for the 108 equations was estimated using a five minute timestep with daily output from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006. Irrelevant parameters were removed for several reasons; parameters were not relevant for the \ac{zerod} model (i.e., hydrodynamic parameters), were considered physical constants, or had no effect given initial conditions.  Additionally, FishTank includes six phytoplankton and two zooplankton groups to add complexity in community structure and foodweb dynamics. To remove obvious redundancies, the sensitivity analyses were conducted using only one phytoplankton and one zooplankton group.  The final set that was evaluated included \Sexpr{length(p1z1pars)} parameters that were further grouped into one of six categories based on applicable biogeochemical components of the model: optics ($n = $ \Sexpr{nparms$Optics} parameters), organic matter (\Sexpr{nparms$`Organic Matter`}), phytoplankton (\Sexpr{nparms$Phytoplankton}), temperature (\Sexpr{nparms$Temperature}), and zooplankton (\Sexpr{nparms$Zooplankton}).  A full description of the model parameters is available as an appendix in \citet{Lehrter17}.
